---
title: "Module 11 — Quality Control of Genomic Data"
description: "MAF, call rate and HWE filters. Detection of problematic samples before analysis."
lang: en
---

```{=html}
<div class="module-header">
  <span class="module-badge">MODULE 11 · GENOMICS</span>
  <div class="module-meta"><span>⏱ ~35 min</span><span>Level: Advanced</span></div>
</div>
<div class="module-objectives">
  <h3>What will you learn</h3>
  <ul>
    <li>Why quality control (QC) is a mandatory step before any genomic analysis</li>
    <li>Apply MAF filters, call rate by SNP and by animal</li>
    <li>Use the HWE test to detect SNPs with systematic errors</li>
    <li>Quantify the impact of QC on the dataset</li>
  </ul>
</div>
<div class="module-orientation" aria-label="How to study this module">
  <span>Reading</span>
  <span>Simulation in R</span>
  <span>Interpretation</span>
</div>
```

::: {.module-reading-rhythm role="note" aria-label="Recommended reading pace for the module"}
**Reading pace**

Read first to understand the question, run the script to generate evidence and only then move on to the exercise and quiz.
:::

::: {.module-session-plan role="region" aria-label="Short module session plan"}
::: {.module-session-plan-copy}
**Short session plan**

Use this module in a closed session: enter with a question, produce scripted evidence, and leave with a short technical decision.
:::

::: {.module-session-plan-grid role="list" aria-label="Module study session steps"}
::: {.module-session-plan-item role="listitem"}
**Question**

Define what needs to be clearer before reading.
:::

::: {.module-session-plan-item role="listitem"}
**Evidence**

Rotate or revise the main section in R.
:::

::: {.module-session-plan-item role="listitem"}
**Decision**

Write a technical consequence before leaving.
:::
:::
:::

::: {.module-technical-scan role="region" aria-label="How to read the module's formulas, code and tables"}
::: {.module-technical-scan-copy}
**Read the technical part in three passages**

Don't try to solve formula, code and table at the same time. First locate the concept, then run or follow the script, and finally interpret the output as evidence.
:::

::: {.module-technical-scan-grid role="list" aria-label="Module technical reading passages"}
::: {.module-technical-scan-item role="listitem"}
**Formula**

Identify which terms change the decision.
:::

::: {.module-technical-scan-item role="listitem"}
**Code**

See which parameter was simulated or compared.
:::

::: {.module-technical-scan-item role="listitem"}
**Table**

Read the difference that supports the interpretation.
:::
:::
:::

::: {.module-script-lab role="region" aria-label="Module 11 script lab"}
::: {.module-script-lab-copy}
**Script lab**

Use the script to answer which QC filters remove noise without losing useful signal. The page renders the main snippets; the complete file allows you to adjust thresholds and measure impact on the final set.
:::

::: {.module-script-lab-grid role="list" aria-label="How to reproduce Module 11 script"}
::: {.module-script-lab-item role="listitem"}
**R script**

[Open `modulo11.R`](../../scripts/modulo11.R){.entry-link aria-label="Open the full R script for Module 11"}
:::

::: {.module-script-lab-item role="listitem"}
**Generated output**

[View simulated CSV](../../data/modulo11_simulado.csv){.entry-link aria-label="Open the simulated CSV generated by Module 11"}
:::

::: {.module-script-lab-item role="listitem"}
**What to change**

Change MAF thresholds and call rate to compare SNP retention and residual missings.
:::

::: {.module-script-lab-item role="listitem"}
**What to interpret**

Explain which filter most changes the balance between genomic coverage and reliability.
:::
:::
:::

## Why raw genomic data is not ready for analysis

Genotyping platforms produce high-quality but imperfect data. Cluster errors, DNA extraction problems, difficult-to-genotype loci, and rare variants complicate analyses. Bad data included in GWAS or in the construction of the G matrix can compromise all results.

The standard QC protocol applies filters at two levels: **per SNP** and **per animal**.

::: {.callout-note}
**Core Concept:** QC doesn't "create" better data—it eliminates problematic data. The resulting set is smaller but more reliable. Making mistakes in QC (being too liberal) contaminates the analysis; being too conservative wastes information.
:::

## Standard QC Criteria

| Criterion | Typical threshold | Foundation |
|---|---|---|
| MAF by SNP | > 0.05 | Rare SNPs have low power and higher error rate |
| Call rate by SNP | ≥ 0.95 | Loci with lots of missing data indicate technical problem |
| Call rate per animal | ≥ 0.90 | Degraded DNA compromises the entire sample |
| HWE p-value | > 10⁻⁶ | Strong deviation indicates systematic genotyping error |

## Simulation and application of QC in R

```{r}
#| label: simula-dados-qc
set.seed(111)
n_anim <- 300; n_snp <- 1000

# Gera dados com problemas propositais
p_alelo <- rbeta(n_snp, 2, 2)
p_alelo[1:80] <- runif(80, 0.01, 0.04)  # SNPs raros (MAF < 0.05)

G <- matrix(NA, n_anim, n_snp)
for (j in 1:n_snp) {
  a1 <- rbinom(n_anim,1,p_alelo[j]); a2 <- rbinom(n_anim,1,p_alelo[j])
  G[,j] <- a1 + a2
}
# Insere missings: SNPs com baixo call rate
snp_ruim <- sample(1:n_snp, 60)
for (j in snp_ruim) G[sample(n_anim,80),j] <- NA
# Amostras com baixo call rate
anim_ruim <- sample(1:n_anim, 15)
G[anim_ruim, sample(n_snp,120)] <- NA

cat(sprintf("Dados brutos: %d animais × %d SNPs\n", n_anim, n_snp))
cat(sprintf("Missings totais: %.1f%%\n", 100*mean(is.na(G))))
```

```{r}
#| label: filtro-snp-cr
# 1. Call rate por SNP
snp_cr <- 1 - colMeans(is.na(G))
cat("=== Filtro 1: Call rate por SNP (≥ 0.95) ===\n")
cat(sprintf("SNPs removidos: %d / %d (%.1f%%)\n",
            sum(snp_cr<.95), n_snp, 100*mean(snp_cr<.95)))
print(quantile(snp_cr, c(0,0.05,0.25,0.5,0.75,0.95,1)))
```

```{r}
#| label: filtro-maf
# 2. MAF
p_obs <- colMeans(G, na.rm=TRUE)/2
maf   <- pmin(p_obs, 1-p_obs)
cat("\n=== Filtro 2: MAF (≥ 0.05) ===\n")
cat(sprintf("SNPs removidos: %d / %d (%.1f%%)\n",
            sum(maf<.05,na.rm=TRUE), n_snp, 100*mean(maf<.05,na.rm=TRUE)))
```

```{r}
#| label: filtro-hwe
# 3. Teste HWE por SNP
hwe_p <- numeric(n_snp)
for (j in 1:n_snp) {
  g <- G[!is.na(G[,j]),j]
  if (length(g)<20) { hwe_p[j]<-0; next }
  pj <- mean(g)/2; qj <- 1-pj
  exp_c <- c(pj^2, 2*pj*qj, qj^2)*length(g)
  obs_c <- as.numeric(table(factor(g,levels=0:2)))
  chi2  <- sum((obs_c-exp_c)^2/(exp_c+1e-9))
  hwe_p[j] <- pchisq(chi2, 1, lower.tail=FALSE)
}
cat("\n=== Filtro 3: HWE (p > 1e-6) ===\n")
cat(sprintf("SNPs fora de HWE: %d / %d (%.1f%%)\n",
            sum(hwe_p<1e-6), n_snp, 100*mean(hwe_p<1e-6)))
```

```{r}
#| label: filtro-anim
# 4. Call rate por animal
anim_cr <- 1 - rowMeans(is.na(G))
cat("\n=== Filtro 4: Call rate por animal (≥ 0.90) ===\n")
cat(sprintf("Animais removidos: %d / %d (%.1f%%)\n",
            sum(anim_cr<.90), n_anim, 100*mean(anim_cr<.90)))
```

```{r}
#| label: aplica-qc
# Aplica todos os filtros
keep_snp  <- snp_cr >= .95 & !is.na(maf) & maf >= .05 & hwe_p >= 1e-6
keep_anim <- anim_cr >= .90

G_qc <- G[keep_anim, keep_snp]

cat("\n=== RESUMO DO CONTROLE DE QUALIDADE ===\n")
cat(sprintf("Antes:   %d animais × %d SNPs\n", n_anim, n_snp))
cat(sprintf("Depois:  %d animais × %d SNPs\n", sum(keep_anim), sum(keep_snp)))
cat(sprintf("Animais retidos: %.1f%%   SNPs retidos: %.1f%%\n",
            100*mean(keep_anim), 100*mean(keep_snp)))
cat(sprintf("Missings após QC: %.2f%%\n", 100*mean(is.na(G_qc))))
```

```{r}
#| fig-height: 3.5
#| label: plot-qc
par(mfrow=c(1,2), mar=c(4,4,2,1))
hist(snp_cr, breaks=30, col="#e8f1f8", border="#0d2642",
     main="Call rate por SNP", xlab="Call rate", ylab="Frequência")
abline(v=0.95, col="#dc2626", lty=2, lwd=2)
hist(anim_cr, breaks=20, col="#e8f1f8", border="#0d2642",
     main="Call rate por animal", xlab="Call rate", ylab="Frequência")
abline(v=0.90, col="#dc2626", lty=2, lwd=2)
```

## Interpretation

- The histograms show that most SNPs and animals have a call rate close to 1.0 — low values ​​are clear exceptions, making it easier to define thresholds.
- The MAF filter removed ~8% of SNPs. These are rare SNPs introduced intentionally.
- After QC, the data has ~0% missing — much cleaner for building the G matrix.

::: {.callout-important}
**Interpretation:** QC thresholds are researcher decisions, not absolute rules. In studies of purebreds with high inbreeding, the MAF threshold can be lowered to 0.01. In mixed populations, SNPs that are rare in one breed may be common in another.
:::

::: {.module-evidence-path role="note" aria-label="Module evidence path"}
**Path of evidence**

Before the exercise, connect three points: which question the module opened, which script output supports the answer and which technical decision is clearer.
:::

::: {.module-practice-contract role="note" aria-label="Module practice contract"}
**Minimal evidence for this module**

Before proceeding, compare two QC filters and write which one most changed the final set of SNPs or animals.
:::

## Proposed exercise

1. Change the call rate threshold per SNP to 0.90 (more liberal). How many extra SNPs are kept?
2. Only remove animals with a call rate < 0.80. How does the result change?
3. After QC, check whether the retained SNPs still have reasonable MAF distribution.

::: {.module-study-checkpoint role="note" aria-label="Study checkpoint before the quiz"}
**Before the quiz**

Review the biological question, run the main section in R, and write a sentence linking the result to the selection decision. Then proceed to the quiz with that phrase in mind.
:::

::: {.module-takeaways role="note" aria-label="Module takeaways"}
**What to take away from this module**

- Quality control removes fragile signals before they contaminate genomic analyses.
- Call rate, MAF, HWE and animal retention must be evaluated together, not as isolated filters.
:::

::: {.quiz-container data-module="11"}
:::

::: {.module-after-quiz role="note" aria-label="Next step after the quiz"}
**After the quiz**

If the quiz result does not reflect your interpretation, return to the excerpt in R and the proposed exercise. If it is consistent, move forward by keeping a technical phrase recorded.
:::

::: {.module-close-check role="region" aria-label="Module closing check"}
::: {.module-close-check-copy}
**Before changing pages**

Close the module when the reading, code and quiz support the same conclusion. If any point is left open, go back through the index or search before moving forward.
:::

::: {.module-close-check-grid role="list" aria-label="Module closing criteria"}
::: {.module-close-check-item role="listitem"}
**Reading**

The central question was answered with his words.
:::

::: {.module-close-check-item role="listitem"}
**Code**

The main output was interpreted, not just executed.
:::

::: {.module-close-check-item role="listitem"}
**Quiz**

The result confirms readiness to move forward or revise.
:::
:::
:::

::: {.module-return-note role="note" aria-label="How to return to the flow after this module"}
**Come back with a decision**

Before changing pages, record whether this module led to advancement, revision or consultation. Use this decision to choose the next click in the index.
:::

```{=html}
<nav class="module-nav" aria-label="Navigation between modules">
  <a class="module-nav-card module-nav-index" href="index.html">
    <span>Index</span>
    <strong>All modules</strong>
  </a>
  <a class="module-nav-card module-nav-prev" href="modulo10-introducao-a-genomica-e-marcadores-snp.html">
    <span>Previous · Module 10</span>
    <strong>Genomics and SNP Markers</strong>
  </a>
  <a class="module-nav-card module-nav-next" href="modulo12-matrizes-genomicas-gwas-e-predicao-genomica.html">
    <span>Next · Module 12</span>
    <strong>Genomic Arrays, GWAS and Prediction</strong>
  </a>
</nav>
```
