---
title: "Module 03 — Descriptive Statistics and Data Exploration in R"
description: "Initial diagnosis of phenotypic data: summaries, distributions, outliers and exploration with R base."
lang: en
---

```{=html}
<div class="module-header">
  <span class="module-badge">MODULE 03 · STATISTICS & R</span>
  <div class="module-meta"><span>⏱ ~30 min</span><span>Level: Introductory</span></div>
</div>
<div class="module-objectives">
  <h3>What will you learn</h3>
  <ul>
    <li>Calculate and interpret position and dispersion measures in phenotypic data</li>
    <li>Visualize distributions with histogram and boxplot in base R</li>
    <li>Identify outliers and evaluate distribution symmetry</li>
    <li>Explore data by groups (breed, farm) with<code>tapply</code>and<code>aggregate</code></li>
  </ul>
</div>
<div class="module-orientation" aria-label="How to study this module">
  <span>Reading</span>
  <span>Simulation in R</span>
  <span>Interpretation</span>
</div>
```

::: {.module-phase-start role="note" aria-label="Opening of the data and parameters phase"}
**New stage of the trail**

Here begins the transition from concept to data diagnosis. Before estimating parameters, the priority is to understand the distribution, variation and quality of the observed phenotype.
:::

::: {.module-reading-rhythm role="note" aria-label="Recommended reading pace for the module"}
**Reading pace**

Read first to understand the question, run the script to generate evidence and only then move on to the exercise and quiz.
:::

::: {.module-session-plan role="region" aria-label="Short module session plan"}
::: {.module-session-plan-copy}
**Short session plan**

Use this module in a closed session: enter with a question, produce scripted evidence, and leave with a short technical decision.
:::

::: {.module-session-plan-grid role="list" aria-label="Module study session steps"}
::: {.module-session-plan-item role="listitem"}
**Question**

Define what needs to be clearer before reading.
:::

::: {.module-session-plan-item role="listitem"}
**Evidence**

Rotate or revise the main section in R.
:::

::: {.module-session-plan-item role="listitem"}
**Decision**

Write a technical consequence before leaving.
:::
:::
:::

::: {.module-technical-scan role="region" aria-label="How to read the module's formulas, code and tables"}
::: {.module-technical-scan-copy}
**Read the technical part in three passages**

Don't try to solve formula, code and table at the same time. First locate the concept, then run or follow the script, and finally interpret the output as evidence.
:::

::: {.module-technical-scan-grid role="list" aria-label="Module technical reading passages"}
::: {.module-technical-scan-item role="listitem"}
**Formula**

Identify which terms change the decision.
:::

::: {.module-technical-scan-item role="listitem"}
**Code**

See which parameter was simulated or compared.
:::

::: {.module-technical-scan-item role="listitem"}
**Table**

Read the difference that supports the interpretation.
:::
:::
:::

::: {.module-script-lab role="region" aria-label="Module 03 script lab"}
::: {.module-script-lab-copy}
**Script lab**

Use the script to answer where dispersion, asymmetry, and outliers exist that could skew the genetic decision. The page renders the main snippets; the complete file allows you to reproduce and vary the diagnosis.
:::

::: {.module-script-lab-grid role="list" aria-label="How to reproduce Module 03 script"}
::: {.module-script-lab-item role="listitem"}
**R script**

[Open `modulo03.R`](../../scripts/modulo03.R){.entry-link aria-label="Open the full R script for Module 03"}
:::

::: {.module-script-lab-item role="listitem"}
**Generated output**

[View simulated CSV](../../data/modulo03_simulado.csv){.entry-link aria-label="Open the simulated CSV generated by Module 03"}
:::

::: {.module-script-lab-item role="listitem"}
**What to change**

Change the outlier threshold (|z| > 3 to |z| > 2.5) and compare how many animals enter review.
:::

::: {.module-script-lab-item role="listitem"}
**What to interpret**

Explain whether distribution by breed and farm supports fixed effects adjustment prior to genetic evaluation.
:::
:::
:::

## Before estimating any parameters, explore the data

In animal breeding, phenotypic data rarely arrives ready for analysis. Typing errors, measurements outside the biologically plausible range, skewed distributions—all of these distort estimates of means, variances, and heritability. Descriptive analysis is not a bureaucratic step: it is the first scientific diagnosis of the data.

::: {.callout-note}
**Central concept:** Descriptive statistics does not answer "what is the genetic value of the animal?", but rather "does the data have sufficient quality to answer this question?".
:::

## Fundamental measures

| Measure | R Formula/Function | Interpretation in improvement |
|---|---|---|
| Average | `mean(x)` | Population phenotypic average — target to be moved |
| Variance | `var(x)` | Direct input for h² = VA/VP |
| DP | `sd(x)` | Variation scale; enters σ_A |
| CV | `sd/mean × 100` | Compares variability between characteristics |
| Asymmetry | `(mean-median)/sd` | Detects departure from normality |

## Simulation and exploration in R

```{r}
#| label: simula-dados
set.seed(103)
n <- 400

raca    <- sample(c("Nelore","Angus","Cruzado"), n, TRUE, prob=c(.45,.25,.30))
fazenda <- sample(paste0("F", 1:5), n, TRUE)
idade   <- round(runif(n, 180, 730))

mu_raca <- c(Nelore=280, Angus=320, Cruzado=300)
peso    <- mu_raca[raca] + 0.08*(idade - 450) + rnorm(n,0,sqrt(120)) + rnorm(n,0,sqrt(400))
peso    <- pmax(peso, 150)

dados <- data.frame(raca=raca, fazenda=fazenda, idade_dias=idade, peso_kg=round(peso,1))
```

```{r}
#| label: descritivas-gerais
cat("=== Resumo do conjunto de dados ===\n")
cat(sprintf("n = %d animais\n\n", nrow(dados)))
print(summary(dados$peso_kg))

cat(sprintf("\nCoef. Variação: %.1f%%\n", 100*sd(dados$peso_kg)/mean(dados$peso_kg)))
cat(sprintf("Assimetria aprox.: %.4f\n",
            (mean(dados$peso_kg) - median(dados$peso_kg)) / sd(dados$peso_kg)))
```

```{r}
#| label: descritivas-por-grupo
cat("=== Médias por raça ===\n")
stats_raca <- aggregate(peso_kg ~ raca, data=dados,
  FUN=function(x) c(n=length(x), media=round(mean(x),1),
                    dp=round(sd(x),1), cv=round(100*sd(x)/mean(x),1)))
print(stats_raca)

cat("\n=== Correlação peso × idade (covariável) ===\n")
cat(sprintf("r = %.4f\n", cor(dados$peso_kg, dados$idade_dias)))
```

```{r}
#| label: outliers
cat("=== Detecção de outliers (|z| > 3) ===\n")
z_scores <- scale(dados$peso_kg)
outliers <- dados[abs(z_scores) > 3, ]
cat(sprintf("Animais suspeitos: %d de %d (%.1f%%)\n",
            nrow(outliers), nrow(dados), 100*nrow(outliers)/nrow(dados)))
if (nrow(outliers) > 0) {
  cat("Pesos dos outliers:\n")
  print(outliers$peso_kg)
}
```

```{r}
#| label: visualizacao
#| fig-height: 4
par(mfrow=c(1,2), mar=c(4,4,2,1))

hist(dados$peso_kg, breaks=25, col="#e8f1f8", border="#0d2642",
     main="Distribuição do peso", xlab="Peso (kg)", ylab="Frequência")
abline(v=mean(dados$peso_kg), col="#0d2642", lwd=2, lty=2)
legend("topright", "Média", lty=2, col="#0d2642", bty="n")

boxplot(peso_kg ~ raca, data=dados, col=c("#4a9fd1","#0d2642","#e8f1f8"),
        main="Peso por raça", xlab="Raça", ylab="Peso (kg)")
```

## Interpretation

- The histogram shows that the weights follow an **approximately normal** distribution, with slight asymmetry — expected in this type of data.
- The boxplot by race shows differences in means between groups. This breed effect **must be corrected** before estimating breeding values.
- The detected outliers (|z| > 3) deserve field review: they could be recording errors or animals with special conditions.
- The positive weight × age correlation confirms the need to include age as a **covariate** in the evaluation model.

::: {.callout-warning}
**Attention:** Never estimate heritability from data without first checking the distribution of residuals. Outliers inflate the residual variance and deflate h².
:::

::: {.module-evidence-path role="note" aria-label="Module evidence path"}
**Path of evidence**

Before the exercise, connect three points: which question the module opened, which script output supports the answer and which technical decision is clearer.
:::

::: {.module-practice-contract role="note" aria-label="Module practice contract"}
**Minimal evidence for this module**

Before proceeding, identify a variable with greater dispersion and write down what check you would do before using this data in a selection decision.
:::

## Proposed exercise

1. Use `tapply(dados$peso_kg, dados$fazenda, mean)` to check for farm effect. What do you observe?
2. Create a subset with just `raca == "Nelore"` and recalculate the CV. Was it bigger or smaller than general? Why?
3. Transform the weight to a logarithmic scale (`log(peso)`) and replot the histogram. Does the asymmetry change?

::: {.module-study-checkpoint role="note" aria-label="Study checkpoint before the quiz"}
**Before the quiz**

Review the biological question, run the main section in R, and write a sentence linking the result to the selection decision. Then proceed to the quiz with that phrase in mind.
:::

::: {.module-takeaways role="note" aria-label="Module takeaways"}
**What to take away from this module**

- Descriptive statistics is screening: it reveals scale, dispersion, groups, and suspect values ​​before the model.
- Graphs and summaries need to answer whether the data supports the initial biological question.
:::

::: {.quiz-container data-module="03"}
:::

::: {.module-after-quiz role="note" aria-label="Next step after the quiz"}
**After the quiz**

If the quiz result does not reflect your interpretation, return to the excerpt in R and the proposed exercise. If it is consistent, move forward by keeping a technical phrase recorded.
:::

::: {.module-close-check role="region" aria-label="Module closing check"}
::: {.module-close-check-copy}
**Before changing pages**

Close the module when the reading, code and quiz support the same conclusion. If any point is left open, go back through the index or search before moving forward.
:::

::: {.module-close-check-grid role="list" aria-label="Module closing criteria"}
::: {.module-close-check-item role="listitem"}
**Reading**

The central question was answered with his words.
:::

::: {.module-close-check-item role="listitem"}
**Code**

The main output was interpreted, not just executed.
:::

::: {.module-close-check-item role="listitem"}
**Quiz**

The result confirms readiness to move forward or revise.
:::
:::
:::

::: {.module-return-note role="note" aria-label="How to return to the flow after this module"}
**Come back with a decision**

Before changing pages, record whether this module led to advancement, revision or consultation. Use this decision to choose the next click in the index.
:::

```{=html}
<nav class="module-nav" aria-label="Navigation between modules">
  <a class="module-nav-card module-nav-index" href="index.html">
    <span>Index</span>
    <strong>All modules</strong>
  </a>
  <a class="module-nav-card module-nav-prev" href="modulo02-bases-da-genetica-quantitativa.html">
    <span>Previous · Module 02</span>
    <strong>Bases of Quantitative Genetics</strong>
  </a>
  <a class="module-nav-card module-nav-next" href="modulo04-medias-variancias-e-componentes-de-variancia.html">
    <span>Next · Module 04</span>
    <strong>Means, Variances and Components</strong>
  </a>
</nav>
```
