---
title: "Module 05 — Heritability and Repeatability"
description: "Concepts, estimation and interpretation of h² and repeatability. Practical applications in breeding programs."
lang: en
---

```{=html}
<div class="module-header">
  <span class="module-badge">MODULE 05 · GENETIC PARAMETERS</span>
  <div class="module-meta"><span>⏱ ~35 min</span><span>Level: Intermediate</span></div>
</div>
<div class="module-objectives">
  <h3>What will you learn</h3>
  <ul>
    <li>What h² really means — and what it doesn't mean</li>
    <li>Estimate h² by parent–child regression and interpret it correctly</li>
    <li>Differentiating heritability from repeatability and when each matters</li>
    <li>Calculate the optimal number of measurements per animal given r</li>
  </ul>
</div>
<div class="module-orientation" aria-label="How to study this module">
  <span>Reading</span>
  <span>Simulation in R</span>
  <span>Interpretation</span>
</div>
```

::: {.module-reading-rhythm role="note" aria-label="Recommended reading pace for the module"}
**Reading pace**

Read first to understand the question, run the script to generate evidence and only then move on to the exercise and quiz.
:::

::: {.module-session-plan role="region" aria-label="Short module session plan"}
::: {.module-session-plan-copy}
**Short session plan**

Use this module in a closed session: enter with a question, produce scripted evidence, and leave with a short technical decision.
:::

::: {.module-session-plan-grid role="list" aria-label="Module study session steps"}
::: {.module-session-plan-item role="listitem"}
**Question**

Define what needs to be clearer before reading.
:::

::: {.module-session-plan-item role="listitem"}
**Evidence**

Rotate or revise the main section in R.
:::

::: {.module-session-plan-item role="listitem"}
**Decision**

Write a technical consequence before leaving.
:::
:::
:::

::: {.module-technical-scan role="region" aria-label="How to read the module's formulas, code and tables"}
::: {.module-technical-scan-copy}
**Read the technical part in three passages**

Don't try to solve formula, code and table at the same time. First locate the concept, then run or follow the script, and finally interpret the output as evidence.
:::

::: {.module-technical-scan-grid role="list" aria-label="Module technical reading passages"}
::: {.module-technical-scan-item role="listitem"}
**Formula**

Identify which terms change the decision.
:::

::: {.module-technical-scan-item role="listitem"}
**Code**

See which parameter was simulated or compared.
:::

::: {.module-technical-scan-item role="listitem"}
**Table**

Read the difference that supports the interpretation.
:::
:::
:::

::: {.module-script-lab role="region" aria-label="Module 05 script lab"}
::: {.module-script-lab-copy}
**Script lab**

Use the script to answer how h² and repeatability change the decision between direct selection and repeated measurements. The page renders the main snippets; The complete file allows you to test collection scenarios.
:::

::: {.module-script-lab-grid role="list" aria-label="How to reproduce Module 05 script"}
::: {.module-script-lab-item role="listitem"}
**R script**

[Open `modulo05.R`](../../scripts/modulo05.R){.entry-link aria-label="Open the full R script for Module 05"}
:::

::: {.module-script-lab-item role="listitem"}
**Generated output**

[View simulated CSV](../../data/modulo05_simulado.csv){.entry-link aria-label="Open the simulated CSV generated by Module 05"}
:::

::: {.module-script-lab-item role="listitem"}
**What to change**

Test values ​​of `n_med` and compare marginal information gain in each new measurement.
:::

::: {.module-script-lab-item role="listitem"}
**What to interpret**

Explain when repeat collection improves decision more than increasing selection intensity.
:::
:::
:::

## What heritability is — and what it isn’t

Heritability (h²) is the most cited and most misinterpreted parameter in animal breeding. It **doesn't** say how much of an animal's phenotype is genetic. It tells what **fraction of the phenotypic variation observed in a population** is attributable to additive genetic differences between animals.

$$h^2 = \frac{V_A}{V_P}$$

::: {.callout-note}
**Central concept:** h² is a property of the **population**, not the individual. The same genome can have different h² in different populations (different environments alter VE) or for different characteristics.
:::

| Feature | typical h² | Implication |
|---|---|---|
| Weight per year | 0.35–0.50 | Responds well to direct selection |
| Milk production | 0.25–0.35 | Respond moderately |
| Pregnancy rate | 0.05–0.15 | Ineffective direct selection |
| Conformation | 0.30–0.45 | Respond reasonably |

## Repeatability: when the animal is measured several times

For characteristics such as milk production (multiple lactations) or weight (multiple weighings), the same animal is measured more than once. **repeatability** (r) measures how much the repeated measurements of an animal agree with each other:

$$r = \frac{V_A + V_D + V_{PE}}{V_P}$$

Where **V_PE** is the permanent environmental variance — effects that persist throughout the animal's life (e.g., an early disease that permanently reduces production).

**Fundamental rule:** r ≥ h² (always — as the numerator of r contains VA plus other positive terms).

```{=html}
<div class="mg-viz" data-mg-viz="h2"></div>
```

## Parent–child regression estimation

```{r}
#| label: regressao-pf
set.seed(105)
n <- 600; va <- 300; ve <- 580; vd <- 80; vpe <- 60

h2  <- va / (va + vd + vpe + ve)
rep <- (va + vd + vpe) / (va + vd + vpe + ve)

# Simula pares pai-filho
a_pai   <- rnorm(n, 0, sqrt(va))
p_pai   <- 250 + a_pai + rnorm(n, 0, sqrt(ve + vd + vpe))
a_filho <- 0.5*a_pai + rnorm(n, 0, sqrt(0.75*va + ve + vd + vpe))
p_filho <- 250 + a_filho + rnorm(n, 0, sqrt(ve + vd + vpe))

reg <- lm(p_filho ~ p_pai)
b   <- coef(reg)[2]

cat("=== Parâmetros teóricos ===\n")
cat(sprintf("h²:           %.4f\n", h2))
cat(sprintf("r (repeat.):  %.4f\n", rep))
cat(sprintf("\n=== Estimativa por regressão pai-filho ===\n"))
cat(sprintf("b observado:  %.4f\n", b))
cat(sprintf("2b (= h²):    %.4f\n", 2*b))
```

```{r}
#| fig-height: 4
#| label: plot-regressao
plot(p_pai, p_filho, pch=20, col=adjustcolor("#0d2642", 0.3),
     xlab="Fenótipo do pai (kg)", ylab="Fenótipo do filho (kg)",
     main=sprintf("Regressão pai–filho  (b = %.3f,  h²est = %.3f)", b, 2*b))
abline(reg, col="#dc2626", lwd=2)
legend("topleft", sprintf("h² teórico = %.3f", h2), bty="n", cex=0.9)
```

```{r}
#| label: repeatability-icc
# Repetibilidade estimada pelo coeficiente de correlação intraclasse (ICC)
n_anim <- 200; n_med <- 4
a_anim  <- rnorm(n_anim, 0, sqrt(va))
pe_anim <- rnorm(n_anim, 0, sqrt(vpe))
medidas <- matrix(NA, n_anim, n_med)
for (m in 1:n_med)
  medidas[, m] <- 250 + a_anim + pe_anim + rnorm(n_anim, 0, sqrt(ve))

# ICC via ANOVA
y_vec <- as.vector(medidas)
gp    <- rep(1:n_anim, times=n_med)
tb    <- anova(lm(y_vec ~ as.factor(gp)))
MS_e  <- tb[1,"Mean Sq"]; MS_d <- tb[2,"Mean Sq"]
r_est <- (MS_e - MS_d) / (MS_e + (n_med-1)*MS_d)

cat("=== Repetibilidade (ICC) ===\n")
cat(sprintf("r estimado: %.4f  (teórico: %.4f)\n", r_est, rep))
```

```{r}
#| label: k-medidas
cat("\n=== Acurácia cumulativa com k medidas ===\n")
cat("(r acumulada = √[k×r / (1+(k-1)r)])\n\n")
for (k in 1:6) {
  r_k <- sqrt(k * rep / (1 + (k-1)*rep))
  ganho <- ifelse(k==1, NA, r_k - sqrt((k-1)*rep/(1+(k-2)*rep)))
  cat(sprintf("k=%d  r_cum=%.4f  ganho marginal: %s\n",
              k, r_k, if(is.na(ganho)) "—" else sprintf("%.4f", ganho)))
}
```

## Interpretation

- The parent–child regression graph shows the slope b ≈ h²/2 ≈ 0.18. High dispersion reflects environmental noise.
- With r = 0.37, **3 measurements** already capture 94% of the information that infinite measurements would capture. The 4th measure onwards has diminishing marginal return.
- The rule of thumb: for features with r < 0,30, vale a pena colher 3–4 medidas por animal; para r > 0.60, 1–2 measurements are sufficient.

::: {.callout-important}
**Interpretation:** low h² does not mean that the characteristic "has no genetics". It means that VE is large in relation to VA in that population — possibly due to high variation in management between farms.
:::

::: {.module-evidence-path role="note" aria-label="Module evidence path"}
**Path of evidence**

Before the exercise, connect three points: which question the module opened, which script output supports the answer and which technical decision is clearer.
:::

::: {.module-practice-contract role="note" aria-label="Module practice contract"}
**Minimal evidence for this module**

Before moving on, write down when it would be worth collecting repeated measurements rather than selecting based on a single observation.
:::

## Proposed exercise

1. Change `n_med` to 2 and recalculate r_est. Does the estimate get worse? Why?
2. Calculate the optimal number of measurements (`k*`) for r = 0.20, 0.40 and 0.70, where the marginal gain < 0.01.
3. If h² = 0.10 and r = 0.45, which variance component must be large? (Hint: V_PE)

::: {.module-study-checkpoint role="note" aria-label="Study checkpoint before the quiz"}
**Before the quiz**

Review the biological question, run the main section in R, and write a sentence linking the result to the selection decision. Then proceed to the quiz with that phrase in mind.
:::

::: {.module-takeaways role="note" aria-label="Module takeaways"}
**What to take away from this module**

- Heritability guides response to selection; Repeatability guides confidence in repeated measurements.
- More records can increase decision certainty, but are not a substitute for good interpretation of the source of variation.
:::

::: {.quiz-container data-module="05"}
:::

::: {.module-after-quiz role="note" aria-label="Next step after the quiz"}
**After the quiz**

If the quiz result does not reflect your interpretation, return to the excerpt in R and the proposed exercise. If it is consistent, move forward by keeping a technical phrase recorded.
:::

::: {.module-close-check role="region" aria-label="Module closing check"}
::: {.module-close-check-copy}
**Before changing pages**

Close the module when the reading, code and quiz support the same conclusion. If any point is left open, go back through the index or search before moving forward.
:::

::: {.module-close-check-grid role="list" aria-label="Module closing criteria"}
::: {.module-close-check-item role="listitem"}
**Reading**

The central question was answered with his words.
:::

::: {.module-close-check-item role="listitem"}
**Code**

The main output was interpreted, not just executed.
:::

::: {.module-close-check-item role="listitem"}
**Quiz**

The result confirms readiness to move forward or revise.
:::
:::
:::

::: {.module-return-note role="note" aria-label="How to return to the flow after this module"}
**Come back with a decision**

Before changing pages, record whether this module led to advancement, revision or consultation. Use this decision to choose the next click in the index.
:::

```{=html}
<nav class="module-nav" aria-label="Navigation between modules">
  <a class="module-nav-card module-nav-index" href="index.html">
    <span>Index</span>
    <strong>All modules</strong>
  </a>
  <a class="module-nav-card module-nav-prev" href="modulo04-medias-variancias-e-componentes-de-variancia.html">
    <span>Previous · Module 04</span>
    <strong>Means, Variances and Components</strong>
  </a>
  <a class="module-nav-card module-nav-next" href="modulo06-correlacoes-geneticas-e-fenotipicas.html">
    <span>Next · Module 06</span>
    <strong>Genetic and Phenotypic Correlations</strong>
  </a>
</nav>
```
