---
title: "Module 08 — BLUP and Genetic Evaluation"
description: "Henderson equations, EBV prediction, accuracy and ranking of selection candidates."
lang: en
---

```{=html}
<div class="module-header">
  <span class="module-badge">MODULE 08 · STATISTICAL MODELS</span>
  <div class="module-meta"><span>⏱ ~45 min</span><span>Level: Advanced</span></div>
</div>
<div class="module-objectives">
  <h3>What will you learn</h3>
  <ul>
    <li>Henderson's Mixed Model Equations (MME)</li>
    <li>Solve the system of equations in R using matrix algebra</li>
    <li>Interpret EBV, DEP and prediction accuracy</li>
    <li>Rank candidates and understand the effect of the number of children</li>
  </ul>
</div>
<div class="module-orientation" aria-label="How to study this module">
  <span>Reading</span>
  <span>Simulation in R</span>
  <span>Interpretation</span>
</div>
```

::: {.module-reading-rhythm role="note" aria-label="Recommended reading pace for the module"}
**Reading pace**

Read first to understand the question, run the script to generate evidence and only then move on to the exercise and quiz.
:::

::: {.module-session-plan role="region" aria-label="Short module session plan"}
::: {.module-session-plan-copy}
**Short session plan**

Use this module in a closed session: enter with a question, produce scripted evidence, and leave with a short technical decision.
:::

::: {.module-session-plan-grid role="list" aria-label="Module study session steps"}
::: {.module-session-plan-item role="listitem"}
**Question**

Define what needs to be clearer before reading.
:::

::: {.module-session-plan-item role="listitem"}
**Evidence**

Rotate or revise the main section in R.
:::

::: {.module-session-plan-item role="listitem"}
**Decision**

Write a technical consequence before leaving.
:::
:::
:::

::: {.module-technical-scan role="region" aria-label="How to read the module's formulas, code and tables"}
::: {.module-technical-scan-copy}
**Read the technical part in three passages**

Don't try to solve formula, code and table at the same time. First locate the concept, then run or follow the script, and finally interpret the output as evidence.
:::

::: {.module-technical-scan-grid role="list" aria-label="Module technical reading passages"}
::: {.module-technical-scan-item role="listitem"}
**Formula**

Identify which terms change the decision.
:::

::: {.module-technical-scan-item role="listitem"}
**Code**

See which parameter was simulated or compared.
:::

::: {.module-technical-scan-item role="listitem"}
**Table**

Read the difference that supports the interpretation.
:::
:::
:::

## The problem Henderson solved

In 1949, C.R. Henderson formulated a system of equations that simultaneously allows:

1. **Estimate** fixed effects (BLUE) — correction for farm, sex, year
2. **Predict** the additive breeding values ​​(BLUP) of each animal

The elegance of the system is that it uses the pedigree directly in the solution, through the inverse of the numerical relationship matrix **A⁻¹**.

## Henderson's equations (MME)

For the `y = Xb + Za + e` animal model:

$$\begin{bmatrix} \mathbf{X'X} & \mathbf{X'Z} \\ \mathbf{Z'X} & \mathbf{Z'Z} + \mathbf{A}^{-1}\lambda \end{bmatrix} \begin{bmatrix} \hat{\mathbf{b}} \\ \hat{\mathbf{a}} \end{bmatrix} = \begin{bmatrix} \mathbf{X'y} \\ \mathbf{Z'y} \end{bmatrix}$$

Where λ = σ²_e / σ²_A. The term **A⁻¹λ** is what "pulls" the EBVs towards the population mean, with strength proportional to the degree of uncertainty about the animal.

::: {.callout-note}
**Central concept:** The inverse of A (A⁻¹) can be calculated directly from the pedigree using the method of Henderson (1976), without ever constructing and inverting A explicitly. This makes BLUP computationally viable for millions of animals.
:::

```{=html}
<div class="mg-viz" data-mg-viz="blup"></div>
```

## Implementation in R

::: {.module-script-lab role="region" aria-label="Module 08 script lab"}
::: {.module-script-lab-copy}
**Script lab**

Use the script to follow the construction of pedigree, matrix A, Henderson equations and ranking by EBV. The practical question is when the genetic ranking changes the decision regarding the raw phenotype.
:::

::: {.module-script-lab-grid role="list" aria-label="How to reproduce Module 08 script"}
::: {.module-script-lab-item role="listitem"}
**R script**

[Open `modulo08.R`](../../scripts/modulo08.R){.entry-link aria-label="Open the full R script for Module 08"}
:::

::: {.module-script-lab-item role="listitem"}
**Generated output**

[View simulated CSV](../../data/modulo08_simulado.csv){.entry-link aria-label="Open the simulated CSV generated by Module 08"}
:::

::: {.module-script-lab-item role="listitem"}
**What to change**

Increase `n_obs` or reduce `h2` and compare accuracy, EBV shrinkage and ranking stability.
:::

::: {.module-script-lab-item role="listitem"}
**What to interpret**

Choose an animal from the ranking and justify whether the selection should follow EBV, phenotype or a combination of the two.
:::
:::
:::

```{r}
#| label: pedigree-a-matrix
set.seed(108)

# Pedigree simplificado: 5 fundadores + 15 animais avaliados
n_fnd <- 5; n_obs <- 15; n_tot <- n_fnd + n_obs

sires <- c(rep(0,n_fnd), sample(1:n_fnd, n_obs, TRUE))
dams  <- c(rep(0,n_fnd), sample(1:n_fnd, n_obs, TRUE))

# Constrói A pelo método tabular (simplificado, sem endogamia)
A <- diag(n_tot)
for (i in (n_fnd+1):n_tot) {
  s <- sires[i]; d <- dams[i]
  if (s>0 && d>0) { A[s,i] <- A[i,s] <- 0.25; A[d,i] <- A[i,d] <- 0.25 }
}
Ainv <- solve(A + 1e-4*diag(n_tot))   # pequena regularização
cat("Dimensão de A:", nrow(A), "×", ncol(A), "\n")
cat("Diagonal de A (primeiros 10):", round(diag(A)[1:10], 3), "\n")
```

```{r}
#| label: simula-fenotipos
# Simula fenótipos (apenas animais não-fundadores observados)
va <- 400; ve <- 600; h2 <- va/(va+ve); lambda <- ve/va

a_true <- rnorm(n_tot, 0, sqrt(va))
obs_id <- (n_fnd+1):n_tot
y      <- 300 + a_true[obs_id] + rnorm(n_obs, 0, sqrt(ve))

cat(sprintf("\nh²=%.3f  λ=%.3f  n observados=%d\n", h2, lambda, n_obs))
```

```{r}
#| label: resolve-mme
# Monta e resolve as MME de Henderson
# Modelo: y = 1μ + Za + e (apenas intercepto como efeito fixo)
X   <- matrix(1, n_obs, 1)
Z   <- matrix(0, n_obs, n_tot)
for (i in seq_along(obs_id)) Z[i, obs_id[i]] <- 1

LHS <- rbind(cbind(t(X)%*%X,          t(X)%*%Z),
             cbind(t(Z)%*%X, t(Z)%*%Z + Ainv*lambda))
RHS <- c(t(X)%*%y, t(Z)%*%y)

sol   <- solve(LHS + 1e-8*diag(nrow(LHS)), RHS)
mu_hat <- sol[1]
a_hat  <- sol[-1]

cat(sprintf("=== Resultados MME ===\n"))
cat(sprintf("μ estimado:     %.2f  (verdadeiro: 300)\n", mu_hat))
cat(sprintf("Cor(EBV, a_verd): %.4f\n", cor(a_hat[obs_id], a_true[obs_id])))
```

```{r}
#| label: ranking-ebv
ebv_df <- data.frame(
  animal       = obs_id,
  sire         = sires[obs_id],
  EBV          = round(a_hat[obs_id], 2),
  a_verdadeiro = round(a_true[obs_id], 2),
  fenotipo     = round(y, 2)
)[order(-a_hat[obs_id]),]

cat("\n=== Top 8 animais por EBV ===\n")
print(head(ebv_df, 8), row.names=FALSE)
```

```{r}
#| fig-height: 4
#| label: plot-ebv
par(mar=c(4,4,2,1))
plot(a_true[obs_id], a_hat[obs_id], pch=20, col=adjustcolor("#0d2642",0.6),
     xlab="a verdadeiro", ylab="EBV (BLUP)",
     main=sprintf("EBV vs. valor genético real  (r=%.3f)", cor(a_hat[obs_id],a_true[obs_id])))
abline(0,1, lty=2, col="#5b6b7d"); abline(lm(a_hat[obs_id]~a_true[obs_id]), col="#dc2626", lwd=2)
legend("topleft", c("Regressão","Linha 45°"), lty=1:2, col=c("#dc2626","#5b6b7d"), bty="n")
```

## Interpretation

- The EBV × a_true correlation is the **assessment accuracy** (r). With only 15 animals evaluated and not many offspring per bull, the accuracy is moderate.
- Note that the EBVs have **less variance** than the true values ​​— the "shrinkage" of BLUP.
- Founder animals (without their own phenotype) receive EBV based **only on offspring** — demonstrating the power of the animal model to use the entire pedigree structure.

::: {.callout-important}
**Interpretation:** DEP (Expected Difference in Progeny) = EBV/2. A bull with EBV = +40 kg has DEP = +20 kg, i.e., on average, his offspring will weigh 20 kg more than offspring of an average bull, under the same conditions.
:::

::: {.module-evidence-path role="note" aria-label="Module evidence path"}
**Path of evidence**

Before the exercise, connect three points: which question the module opened, which script output supports the answer and which technical decision is clearer.
:::

::: {.module-practice-contract role="note" aria-label="Module practice contract"}
**Minimal evidence for this module**

Before moving on, compare phenotype and EBV for a ranking animal and write why the selection decision might change.
:::

## Proposed exercise

1. Increase `n_obs` to 100. Does accuracy improve? Why?
2. Change `h2` to 0.10 (λ too high). Does shrinkage of EBVs increase? Check via `sd(a_hat)`.
3. Select the top 3 animals by EBV. Check if they have the highest true values.

::: {.module-study-checkpoint role="note" aria-label="Study checkpoint before the quiz"}
**Before the quiz**

Review the biological question, run the main section in R, and write a sentence linking the result to the selection decision. Then proceed to the quiz with that phrase in mind.
:::

::: {.module-takeaways role="note" aria-label="Module takeaways"}
**What to take away from this module**

- BLUP combines parentage, phenotype, and variances to predict genetic merit with uncertainty control.
- EBV rankings must be read considering accuracy, data structure and selection objective.
:::

::: {.quiz-container data-module="08"}
:::

::: {.module-after-quiz role="note" aria-label="Next step after the quiz"}
**After the quiz**

If the quiz result does not reflect your interpretation, return to the excerpt in R and the proposed exercise. If it is consistent, move forward by keeping a technical phrase recorded.
:::

::: {.module-close-check role="region" aria-label="Module closing check"}
::: {.module-close-check-copy}
**Before changing pages**

Close the module when the reading, code and quiz support the same conclusion. If any point is left open, go back through the index or search before moving forward.
:::

::: {.module-close-check-grid role="list" aria-label="Module closing criteria"}
::: {.module-close-check-item role="listitem"}
**Reading**

The central question was answered with his words.
:::

::: {.module-close-check-item role="listitem"}
**Code**

The main output was interpreted, not just executed.
:::

::: {.module-close-check-item role="listitem"}
**Quiz**

The result confirms readiness to move forward or revise.
:::
:::
:::

::: {.module-return-note role="note" aria-label="How to return to the flow after this module"}
**Come back with a decision**

Before changing pages, record whether this module led to advancement, revision or consultation. Use this decision to choose the next click in the index.
:::

```{=html}
<nav class="module-nav" aria-label="Navigation between modules">
  <a class="module-nav-card module-nav-index" href="index.html">
    <span>Index</span>
    <strong>All modules</strong>
  </a>
  <a class="module-nav-card module-nav-prev" href="modulo07-modelos-lineares-e-modelos-mistos.html">
    <span>Previous · Module 07</span>
    <strong>Linear and Mixed Models</strong>
  </a>
  <a class="module-nav-card module-nav-next" href="modulo09-estrutura-de-pedigree-e-parentesco.html">
    <span>Next · Module 09</span>
    <strong>Pedigree and Kinship Structure</strong>
  </a>
</nav>
```
