---
title: "Module 10 — Introduction to Genomics and SNP Markers"
description: "SNPs, genotyping platforms, allele frequencies and linkage disequilibrium."
lang: en
---

```{=html}
<div class="module-header">
  <span class="module-badge">MODULE 10 · GENOMICS</span>
  <div class="module-meta"><span>⏱ ~35 min</span><span>Level: Intermediate</span></div>
</div>
<div class="module-objectives">
  <h3>What will you learn</h3>
  <ul>
    <li>What are SNPs and how are they genotyped in populations</li>
    <li>Allele frequency, MAF and 0/1/2 coding</li>
    <li>Linkage disequilibrium (LD) and its importance for GWAS</li>
    <li>Simulate and analyze SNP data with R base</li>
  </ul>
</div>
<div class="module-orientation" aria-label="How to study this module">
  <span>Reading</span>
  <span>Simulation in R</span>
  <span>Interpretation</span>
</div>
```

::: {.module-phase-start role="note" aria-label="Opening of the genomics phase"}
**New stage of the trail**

Here begins the genomic reading of variation. Kinship no longer depends solely on the expected pedigree and starts to use markers observed in DNA.
:::

::: {.module-reading-rhythm role="note" aria-label="Recommended reading pace for the module"}
**Reading pace**

Read first to understand the question, run the script to generate evidence and only then move on to the exercise and quiz.
:::

::: {.module-session-plan role="region" aria-label="Short module session plan"}
::: {.module-session-plan-copy}
**Short session plan**

Use this module in a closed session: enter with a question, produce scripted evidence, and leave with a short technical decision.
:::

::: {.module-session-plan-grid role="list" aria-label="Module study session steps"}
::: {.module-session-plan-item role="listitem"}
**Question**

Define what needs to be clearer before reading.
:::

::: {.module-session-plan-item role="listitem"}
**Evidence**

Rotate or revise the main section in R.
:::

::: {.module-session-plan-item role="listitem"}
**Decision**

Write a technical consequence before leaving.
:::
:::
:::

::: {.module-technical-scan role="region" aria-label="How to read the module's formulas, code and tables"}
::: {.module-technical-scan-copy}
**Read the technical part in three passages**

Don't try to solve formula, code and table at the same time. First locate the concept, then run or follow the script, and finally interpret the output as evidence.
:::

::: {.module-technical-scan-grid role="list" aria-label="Module technical reading passages"}
::: {.module-technical-scan-item role="listitem"}
**Formula**

Identify which terms change the decision.
:::

::: {.module-technical-scan-item role="listitem"}
**Code**

See which parameter was simulated or compared.
:::

::: {.module-technical-scan-item role="listitem"}
**Table**

Read the difference that supports the interpretation.
:::
:::
:::

::: {.module-script-lab role="region" aria-label="Module 10 script lab"}
::: {.module-script-lab-copy}
**Script lab**

Use the script to answer how SNP, MAF, and LD coding qualifies panel prior to genomic analyses. The page renders the main snippets; the complete file allows you to change size and frequency distribution.
:::

::: {.module-script-lab-grid role="list" aria-label="How to reproduce Module 10 script"}
::: {.module-script-lab-item role="listitem"}
**R script**

[Open `modulo10.R`](../../scripts/modulo10.R){.entry-link aria-label="Open the full R script for Module 10"}
:::

::: {.module-script-lab-item role="listitem"}
**Generated output**

[View simulated CSV](../../data/modulo10_simulado.csv){.entry-link aria-label="Open the simulated CSV generated by Module 10"}
:::

::: {.module-script-lab-item role="listitem"}
**What to change**

Change the distribution of `p_alelo` and compare the proportion of SNPs with low MAF and adjacent r².
:::

::: {.module-script-lab-item role="listitem"}
**What to interpret**

Explain whether the generated panel is suitable for GWAS without inflating noise due to rare SNP.
:::
:::
:::

## From pedigree to DNA

Previous modules used pedigree to infer parentage. But the pedigree says what we **expect** on average — a child receives, on average, 50% of the father's genes, but the real value can vary from 40% to 60% by meiotic chance. With **SNP molecular markers**, we measure what the animal **actually inherited**, locus by locus.

A **SNP (Single Nucleotide Polymorphism)** is a position in the genome where there is variation between individuals in a single nitrogenous base (e.g.: A/G, C/T). Commercial platforms such as BovineSNP50 genotype ~50,000 SNPs; BovineHD genotypes ~777,000.

::: {.callout-note}
**Core concept:** The standard coding for SNPs is 0, 1, 2 — the number of copies of the alternative allele. 0 = AA (reference homozygote), 1 = Aa (heterozygote), 2 = aa (alternative homozygote).
:::

## Allele frequencies and MAF

For each SNP j in the population:

$$p_j = \frac{\text{número de alelos alternativos}}{2n}$$

$$MAF_j = \min(p_j, 1-p_j)$$

SNPs with very low MAF (< 0.05) are uninformative and prone to genotyping errors — generally removed in quality control.

## Simulation in R

```{r}
#| label: simula-snp
set.seed(110)
n_anim <- 200; n_snp <- 500

# Frequências alélicas (distribuição Beta para variedade realista)
p_alelo <- rbeta(n_snp, 2, 2)
p_alelo <- pmax(0.02, pmin(0.98, p_alelo))

# Gera matriz de genótipos 0/1/2
G_mat <- matrix(NA, n_anim, n_snp)
for (j in 1:n_snp) {
  a1 <- rbinom(n_anim, 1, p_alelo[j])
  a2 <- rbinom(n_anim, 1, p_alelo[j])
  G_mat[, j] <- a1 + a2
}
cat(sprintf("Matriz de genótipos: %d animais × %d SNPs\n", n_anim, n_snp))
cat(sprintf("Primeiros 3 animais, 8 SNPs:\n"))
print(G_mat[1:3, 1:8])
```

```{r}
#| label: frequencias
p_obs <- colMeans(G_mat) / 2
maf   <- pmin(p_obs, 1 - p_obs)

cat("\n=== Distribuição de MAF ===\n")
print(summary(maf))
cat(sprintf("SNPs com MAF < 0.05: %d (%.1f%%)\n", sum(maf<.05), 100*mean(maf<.05)))
cat(sprintf("SNPs com MAF > 0.40: %d (%.1f%%)\n", sum(maf>.40), 100*mean(maf>.40)))
```

```{r}
#| fig-height: 3.5
#| label: plot-maf
hist(maf, breaks=30, col="#e8f1f8", border="#0d2642",
     main="Distribuição de MAF (n=500 SNPs)", xlab="Minor Allele Frequency", ylab="Frequência")
abline(v=0.05, col="#dc2626", lty=2, lwd=2)
legend("topright", "Limiar MAF=0.05", lty=2, col="#dc2626", bty="n")
```

```{r}
#| label: hwe-test
# Teste de Equilíbrio de Hardy-Weinberg para o SNP 1
j     <- 1
p1    <- p_obs[j]; q1 <- 1 - p1
obs   <- table(factor(G_mat[,j], levels=0:2))
exp   <- c(p1^2, 2*p1*q1, q1^2) * n_anim

cat("=== Teste HWE — SNP 1 ===\n")
cat(sprintf("p observada: %.4f\n", p1))
cat(sprintf("Genótipos observados:  AA=%d  Aa=%d  aa=%d\n", obs[1],obs[2],obs[3]))
cat(sprintf("Genótipos esperados:   AA=%.1f Aa=%.1f aa=%.1f\n", exp[1],exp[2],exp[3]))
chi2 <- sum((obs - exp)^2 / exp)
pval <- pchisq(chi2, df=1, lower.tail=FALSE)
cat(sprintf("χ²=%.4f  p-valor=%.4f  %s\n", chi2, pval,
            if(pval>0.05) "→ HWE não rejeitado" else "→ FORA do HWE"))
```

```{r}
#| label: ld
# Desequilíbrio de ligação entre SNPs adjacentes
r2_adj <- sapply(1:(n_snp-1), function(j) cor(G_mat[,j], G_mat[,j+1])^2)
cat("\n=== Desequilíbrio de ligação entre SNPs adjacentes ===\n")
cat(sprintf("r² médio:   %.4f\n", mean(r2_adj)))
cat(sprintf("r² mediano: %.4f\n", median(r2_adj)))
cat(sprintf("r² máximo:  %.4f\n", max(r2_adj)))
cat("\n(r² alto = SNPs em desequilíbrio = um 'representa' o outro no GWAS)\n")
```

## Interpretation

- The Bell-shaped MAF distribution with mode ~0.30–0.35 is typical of well-designed platforms: few rare variants, many moderately common.
- The HWE test verifies whether the frequencies of AA, Aa and aa genotypes correspond to what is expected in panmixia. Systematic deviations indicate genotyping errors.
- The LD between adjacent SNPs is low (~0.03) in this random position simulation. In the real genome, nearby SNPs have much higher LD, forming **haplotype blocks**.

::: {.callout-tip}
**R Tip:** The `snpStats` (Bioconductor) package is the reference for analyzing SNP data in R. For large data, `PLINK` and `GCTA` are the default command-line tools.
:::

::: {.module-evidence-path role="note" aria-label="Module evidence path"}
**Path of evidence**

Before the exercise, connect three points: which question the module opened, which script output supports the answer and which technical decision is clearer.
:::

::: {.module-practice-contract role="note" aria-label="Module practice contract"}
**Minimal evidence for this module**

Before proceeding, choose a quality criterion for SNPs and write down what risk it reduces before a genomic analysis.
:::

## Proposed exercise

1. Calculate the expected genotype distribution for the SNP with the highest MAF. Compare with those observed.
2. Calculate r² between SNPs separated by 1, 5 and 10 positions. Does LD decay with distance?
3. How does sample size affect the accuracy of estimating p? Recalculate p_obs with only 20 animals.

::: {.module-study-checkpoint role="note" aria-label="Study checkpoint before the quiz"}
**Before the quiz**

Review the biological question, run the main section in R, and write a sentence linking the result to the selection decision. Then proceed to the quiz with that phrase in mind.
:::

::: {.module-takeaways role="note" aria-label="Module takeaways"}
**What to take away from this module**

- SNPs summarize genomic variation into markers that can be counted, filtered and compared.
- Allele frequency, balance and linkage disequilibrium indicate quality and structure of the genomic panel.
:::

::: {.quiz-container data-module="10"}
:::

::: {.module-after-quiz role="note" aria-label="Next step after the quiz"}
**After the quiz**

If the quiz result does not reflect your interpretation, return to the excerpt in R and the proposed exercise. If it is consistent, move forward by keeping a technical phrase recorded.
:::

::: {.module-close-check role="region" aria-label="Module closing check"}
::: {.module-close-check-copy}
**Before changing pages**

Close the module when the reading, code and quiz support the same conclusion. If any point is left open, go back through the index or search before moving forward.
:::

::: {.module-close-check-grid role="list" aria-label="Module closing criteria"}
::: {.module-close-check-item role="listitem"}
**Reading**

The central question was answered with his words.
:::

::: {.module-close-check-item role="listitem"}
**Code**

The main output was interpreted, not just executed.
:::

::: {.module-close-check-item role="listitem"}
**Quiz**

The result confirms readiness to move forward or revise.
:::
:::
:::

::: {.module-return-note role="note" aria-label="How to return to the flow after this module"}
**Come back with a decision**

Before changing pages, record whether this module led to advancement, revision or consultation. Use this decision to choose the next click in the index.
:::

```{=html}
<nav class="module-nav" aria-label="Navigation between modules">
  <a class="module-nav-card module-nav-index" href="index.html">
    <span>Index</span>
    <strong>All modules</strong>
  </a>
  <a class="module-nav-card module-nav-prev" href="modulo09-estrutura-de-pedigree-e-parentesco.html">
    <span>Previous · Module 09</span>
    <strong>Pedigree and Kinship Structure</strong>
  </a>
  <a class="module-nav-card module-nav-next" href="modulo11-controle-de-qualidade-de-dados-genomicos.html">
    <span>Next · Module 11</span>
    <strong>Genomic Quality Control</strong>
  </a>
</nav>
```
