---
title: "Module 02 — Bases of Quantitative Genetics"
description: "The equation P = G + E, decomposition of genetic value and the inheritance of quantitative traits."
lang: en
---

```{=html}
<div class="module-header">
  <span class="module-badge">MODULE 02 · FUNDAMENTALS</span>
  <div class="module-meta"><span>⏱ ~30 min</span><span>Level: Introductory</span></div>
</div>
<div class="module-objectives">
  <h3>What will you learn</h3>
  <ul>
    <li>Phenotype breakdown: P = G + E</li>
    <li>Genotypic value components: A (additive), D (dominance), I (epistasis)</li>
    <li>What is enhancement value and why does it only inherit the additive half</li>
    <li>Estimate h² by parent–child regression</li>
  </ul>
</div>
<div class="module-orientation" aria-label="How to study this module">
  <span>Reading</span>
  <span>Simulation in R</span>
  <span>Interpretation</span>
</div>
```

::: {.module-reading-rhythm role="note" aria-label="Recommended reading pace for the module"}
**Reading pace**

Read first to understand the question, run the script to generate evidence and only then move on to the exercise and quiz.
:::

::: {.module-session-plan role="region" aria-label="Short module session plan"}
::: {.module-session-plan-copy}
**Short session plan**

Use this module in a closed session: enter with a question, produce scripted evidence, and leave with a short technical decision.
:::

::: {.module-session-plan-grid role="list" aria-label="Module study session steps"}
::: {.module-session-plan-item role="listitem"}
**Question**

Define what needs to be clearer before reading.
:::

::: {.module-session-plan-item role="listitem"}
**Evidence**

Rotate or revise the main section in R.
:::

::: {.module-session-plan-item role="listitem"}
**Decision**

Write a technical consequence before leaving.
:::
:::
:::

::: {.module-technical-scan role="region" aria-label="How to read the module's formulas, code and tables"}
::: {.module-technical-scan-copy}
**Read the technical part in three passages**

Don't try to solve formula, code and table at the same time. First locate the concept, then run or follow the script, and finally interpret the output as evidence.
:::

::: {.module-technical-scan-grid role="list" aria-label="Module technical reading passages"}
::: {.module-technical-scan-item role="listitem"}
**Formula**

Identify which terms change the decision.
:::

::: {.module-technical-scan-item role="listitem"}
**Code**

See which parameter was simulated or compared.
:::

::: {.module-technical-scan-item role="listitem"}
**Table**

Read the difference that supports the interpretation.
:::
:::
:::

## Why do zootechnical characteristics not follow Mendel?

Weight, milk production, prolificacy — these characteristics vary continuously between animals. There is no "heavy" or "light", there is a distribution. This happens because **many loci** contribute simultaneously, each with a small effect, and the **environment** superimposes noise on this genetic sum. The result is the normal curve that we observe in zootechnical phenotypes.

The fundamental model is:

$$P = G + E$$

Where **P** is the observed phenotype, **G** the genotypic value and **E** the environmental deviation.

## Breakdown of genotypic value

The genotypic value **G** is not simple. It breaks down into:

$$G = A + D + I$$

| Component | Name | Broadcast |
|---|---|---|
| **A** | Additive value (breeding value) | Half for each child |
| **D** | Dominance deviation | Cannot be transmitted (recombination) |
| **I** | Interaction deviation (epistasis) | Cannot be transmitted (recombination) |

::: {.callout-note}
**Central concept:** The **improvement value** (A) is the only part of the genotype that is transmitted in a predictable way to the progeny. On average, each child receives **half** the parent's A value — hence DEP = EBV / 2.
:::

The phenotypic variance breaks down in an analogous way:

$$V_P = V_A + V_D + V_I + V_E$$

Heritability in the narrow sense — the most important parameter of improvement — is:

$$h^2 = \frac{V_A}{V_P}$$

```{=html}
<div class="mg-viz" data-mg-viz="pge"></div>
```

## Simulation in R

::: {.module-script-lab role="region" aria-label="Module 02 script lab"}
::: {.module-script-lab-copy}
**Script lab**

Use the script to separate additive variance, dominance, epistasis, and environment. The central evidence is to compare `VA/VP`, the `P × A` correlation, and the parent-child regression.
:::

::: {.module-script-lab-grid role="list" aria-label="How to reproduce Module 02 script"}
::: {.module-script-lab-item role="listitem"}
**R script**

[Open `modulo02.R`](../../scripts/modulo02.R){.entry-link aria-label="Open the full R script for Module 02"}
:::

::: {.module-script-lab-item role="listitem"}
**Generated output**

[View simulated CSV](../../data/modulo02_simulado.csv){.entry-link aria-label="Open the simulated CSV generated by Module 02"}
:::

::: {.module-script-lab-item role="listitem"}
**What to change**

Increase `vd` while keeping `va` constant and observe why heritability drops even with the same additive variance.
:::

::: {.module-script-lab-item role="listitem"}
**What to interpret**

Write a conclusion about why high phenotypic variation is not enough to predict response to selection.
:::
:::
:::

```{r}
#| label: decomposicao-vp
set.seed(102)
n  <- 500

# Definir componentes (valores didáticos em unidades de variância)
va <- 200; vd <- 60; vi <- 20; ve <- 320
vp <- va + vd + vi + ve

cat("=== Decomposição da variância fenotípica ===\n")
comps <- c("VA (aditiva)", "VD (dominância)", "VI (epistasia)", "VE (ambiental)", "VP total")
vals  <- c(va, vd, vi, ve, vp)
for (i in seq_along(comps))
  cat(sprintf("%-20s %6.1f  (%5.1f%%)\n", comps[i], vals[i], 100*vals[i]/vp))
cat(sprintf("\nh² = VA/VP = %.4f\n", va/vp))
```

```{r}
#| label: simulacao-pge
# Simula animais com P = A + D + I + E
a_sim <- rnorm(n, 0, sqrt(va))
d_sim <- rnorm(n, 0, sqrt(vd))
i_sim <- rnorm(n, 0, sqrt(vi))
e_sim <- rnorm(n, 0, sqrt(ve))
g_sim <- a_sim + d_sim + i_sim
p_sim <- 100 + g_sim + e_sim

cat("=== Variâncias observadas na simulação ===\n")
cat(sprintf("Var(P):         %6.2f  (esperado: %6.1f)\n", var(p_sim), vp))
cat(sprintf("Var(G):         %6.2f  (esperado: %6.1f)\n", var(g_sim), va+vd+vi))
cat(sprintf("Var(A):         %6.2f  (esperado: %6.1f)\n", var(a_sim), va))
cat(sprintf("Cor(P, A):      %6.4f  (≈ h = √h²)\n", cor(p_sim, a_sim)))
cat(sprintf("h² obs (cor²):  %6.4f\n", cor(p_sim, a_sim)^2))
cat(sprintf("h² teórico:     %6.4f\n", va/vp))
```

```{r}
#| label: regressao-pai-filho
# Regressão filho ~ pai: o coeficiente b estima h²/2
a_pai   <- rnorm(n, 0, sqrt(va))
p_pai   <- 100 + a_pai + rnorm(n, 0, sqrt(ve + vd + vi))
a_filho <- 0.5 * a_pai + rnorm(n, 0, sqrt(0.75 * va + ve + vd))
p_filho <- 100 + a_filho + rnorm(n, 0, sqrt(ve + vd + vi))

b   <- coef(lm(p_filho ~ p_pai))[2]
cat("=== Regressão filho ~ pai ===\n")
cat(sprintf("b (coef. regressão): %.4f\n", b))
cat(sprintf("h² estimado (2×b):   %.4f\n", 2*b))
cat(sprintf("h² teórico:          %.4f\n", va/vp))
```

## Interpretation

- The simulation shows that **only VA contributes** to the correlation between parents and children. VD and VI "disappear" upon recombination.
- The parent–child regression yields b ≈ h²/2 ≈ 0.20 — consistent with the theoretical 0.19. Variations between simulations are expected (finite sampling).
- The h² observed via `cor(P, A)²` converges to the theoretical one with large n.

::: {.callout-important}
**Interpretation:** h² = 0.31 does not mean that "31% of each animal's phenotype is genetic". It means that 31% of the **variation** observed in the population is due to additive genetic differences between animals.
:::

::: {.module-evidence-path role="note" aria-label="Module evidence path"}
**Path of evidence**

Before the exercise, connect three points: which question the module opened, which script output supports the answer and which technical decision is clearer.
:::

::: {.module-practice-contract role="note" aria-label="Module practice contract"}
**Minimal evidence for this module**

Before proceeding, explain in one sentence why high phenotypic variance does not guarantee a response to selection if the additive fraction is low.
:::

## Proposed exercise

1. Increase `vd` to 150 and recalculate `h²`. How did h² change even with constant VA?
2. Simulate what happens with the parent–child regression when `n <- 50`. Compare with `n <- 2000`.
3. What would be the h² if `vi = 0` and `vd = 0`? What does this mean biologically?

::: {.module-study-checkpoint role="note" aria-label="Study checkpoint before the quiz"}
**Before the quiz**

Review the biological question, run the main section in R, and write a sentence linking the result to the selection decision. Then proceed to the quiz with that phrase in mind.
:::

::: {.module-takeaways role="note" aria-label="Module takeaways"}
**What to take away from this module**

- The observed phenotype combines genetic and environmental components, not just genetic value.
- Heritability summarizes the additive fraction useful for selection within a defined population and environment.
:::

::: {.quiz-container data-module="02"}
:::

::: {.module-after-quiz role="note" aria-label="Next step after the quiz"}
**After the quiz**

If the quiz result does not reflect your interpretation, return to the excerpt in R and the proposed exercise. If it is consistent, move forward by keeping a technical phrase recorded.
:::

::: {.module-phase-note role="note" aria-label="Transition to the data and parameters phase"}
**Phase transition**

You have completed the conceptual basis of the track. From the next module onwards, the question stops being just what genetic variation is and becomes how to observe, summarize and interpret simulated data.
:::

::: {.module-close-check role="region" aria-label="Module closing check"}
::: {.module-close-check-copy}
**Before changing pages**

Close the module when the reading, code and quiz support the same conclusion. If any point is left open, go back through the index or search before moving forward.
:::

::: {.module-close-check-grid role="list" aria-label="Module closing criteria"}
::: {.module-close-check-item role="listitem"}
**Reading**

The central question was answered with his words.
:::

::: {.module-close-check-item role="listitem"}
**Code**

The main output was interpreted, not just executed.
:::

::: {.module-close-check-item role="listitem"}
**Quiz**

The result confirms readiness to move forward or revise.
:::
:::
:::

::: {.module-return-note role="note" aria-label="How to return to the flow after this module"}
**Come back with a decision**

Before changing pages, record whether this module led to advancement, revision or consultation. Use this decision to choose the next click in the index.
:::

```{=html}
<nav class="module-nav" aria-label="Navigation between modules">
  <a class="module-nav-card module-nav-index" href="index.html">
    <span>Index</span>
    <strong>All modules</strong>
  </a>
  <a class="module-nav-card module-nav-prev" href="modulo01-introducao-ao-melhoramento-animal.html">
    <span>Previous · Module 01</span>
    <strong>Introduction to Animal Improvement</strong>
  </a>
  <a class="module-nav-card module-nav-next" href="modulo03-estatistica-descritiva-e-exploracao-de-dados-no-r.html">
    <span>Next · Module 03</span>
    <strong>Descriptive Statistics and Exploration in R</strong>
  </a>
</nav>
```
